A Dearth of Young Stars and Planets in the Kepler Field

### 2024-05-22 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The results include \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  This is far fewer than expected under the classic assumption of a uniform
  star formation history: the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star
  formation history has been previously reported, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).  The
    rotation periods are primarily drawn from
    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
    suitable for gyrochronology (black points), based on factors
    including their temperatures, surface gravities, non-binarity, and
    proximity to the main sequence (see Section~\ref{subsec:flags}).
    Some are ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue
    points; Section~\ref{subsec:plflags}).  Surface gravities and
    effective temperatures were derived photometrically by
    \citet{Berger_2020a_catalog}.
    MIST isochrones spanning 3--6\,Gyr in 1\,Gyr increments 
    are shown in the lower panel \citep{Choi_2016} .
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion becomes much larger
    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added concerns of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
	the method was as follows.
The orange line in Figure~\ref{fig:stellarprops} is	defined as
the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
We defined $f_0$ by fitting an
$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
and $T_{\rm eff}\in[3700,5900]$\,K, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  
By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
However for F and G dwarfs, the resulting locus
allowed for stars that were `too evolved''; the density of KOIs 
was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
We therefore visually selected KOIs that appeared
to be near the main sequence -- i.e.~ most of the blue
points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
polynomial $f_1$ through a similar BIC-minimization procedure,
and found $N$=3.
The portion of the orange locus from $\approx$5300--6200\,K
is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
$\approx$0.1\,dex above the ``KOI main sequence''.
The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
We emphasize that for stars
surface gravities above the orange locus, we do report rotation based ages;
they are simply flagged as not being near the main sequence.  }
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.
Some care is required since beyond age,
exoplanet demographics also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018}.

>>>>

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  This is far fewer than expected under the classic assumption of a uniform
  star formation history.  We find that this dearth can be attributed to the star formation
  rate in the Kepler field
  dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend is observed both for known planet hosts as well as for the overall
  stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star
  formation history has been previously reported, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.475\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.475\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{{\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (gray points).  The
    rotation periods are primarily drawn from
    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
    suitable for gyrochronology (dark gray points), based on factors
    including their non-binarity and
    proximity to the spectroscopic main sequence (orange boundary; see Section~\ref{subsec:flags}).
    Some are ``confirmed'' or ``candidate'' KOIs
    that meet an additional set of planetary quality criteria (yellow
    points; Section~\ref{subsec:plflags}).  Surface gravities and
    effective temperatures were derived photometrically by
    \citet{Berger_2020a_catalog}.
    Isochrones 
    in the lower panel are from MIST \citep{Choi_2016} .
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.25cm}
	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion becomes much larger
    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added concerns of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
	the method was as follows.
The orange line in Figure~\ref{fig:stellarprops} is	defined as
the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
We defined $f_0$ by fitting an
$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
and $T_{\rm eff}\in[3700,5900]$\,K, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  
By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
However for F and G dwarfs, the resulting locus
allowed for stars that were `too evolved''; the density of KOIs 
was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
We therefore visually selected KOIs that appeared
to be near the main sequence -- i.e.~ most of the yellow
points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
polynomial $f_1$ through a similar BIC-minimization procedure,
which yielded $N$=3.
This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
The portion of the orange locus from $\approx$5300--6200\,K
is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
0.1\,dex above the ``KOI main sequence''.
The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
We emphasize that we do report rotation-based ages for stars
above the orange locus;
they are simply flagged as not being near the main sequence.  }
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.
Some care is required since beyond age,
exoplanet demographics also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018,Miyazaki2023}.
